---
title: "MovieLens Project Submission"
author: "Kirtimay Pendse"
date: "6/9/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This project is part of the HarvardX: PH125.9x Data Science: Capstone course, the final course for the Data Science Professional Certificate. The aim of this project is to create a movie recommendation system using a subset of the MovieLens dataset. The algorithm was developed using a training set (referred to as the 'edx' set) and was tested on the 'validation' set; then, the random mean squared error (RMSE) was calculated to evaluate the proximity of the predictions generated by the algorithm to the true values in the validation set. This report is split into four sections: first, the objective and motivation behind the project is highlighted, then exploratory data analysis is conducted, following which the modeling approach to develop the predicted movie rating algorithm is presented. Finally, the modeling resutls are presented along with a discussion on the algorithm's performance and its limitations.

## Introduction

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
