---
title: "MovieLens Project Submission"
author: "Kirtimay Pendse"
date: "6/9/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This project is part of the HarvardX: PH125.9x Data Science: Capstone course, the final course for the Data Science Professional Certificate. The aim of this project is to create a movie recommendation system using a subset of the MovieLens dataset. The algorithm was developed using a training set (referred to as the 'edx' set) and was tested on the 'validation' set; then, the random mean squared error (RMSE) was calculated to evaluate the proximity of the predictions generated by the algorithm to the true values in the validation set. This report is split into four sections: first, the objective and motivation behind the project is highlighted, then exploratory data analysis is conducted, following which the modeling approach to develop the predicted movie rating algorithm is presented. Finally, the modeling resutls are presented along with a discussion on the algorithm's performance and its limitations.

## Introduction

The MovieLens dataset is compiled by the GroupLens research group at the University of Minnesota. The dataset contains over 20 million ratings for more than 27,000 movies with around 140,000 users. The overall aim of this project is to develop an algorithm that can predict user ratings (ranging from 0.5 to 5) in the validation set using a subset of the MovieLens dataset provided by the edX staff (the 'edx' dataset).

The algorithm's performance is evaluated using the RMSE, a measure of the difference between the model's predicted values and the observed values which can be written as the following function: 
$$ RMSE = \sqrt{\frac{1}{N}\displaystyle\sum_{u,i} (\hat{y}_{u,i}-y_{u,i})^{2}} $$

```{r RMSE_function1, echo = FALSE}
RMSE <- function(predicted_ratings, true_ratings){
  sqrt(mean((predicted_ratings - true_ratings)^2))
}
```

Multiple models were created and compared using their resulting RMSE values.

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
